Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-22 text/ms.tex
Edits to main ms

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, and
quantitative evidence for a limited number of systems, but systematic evidence
has been elusive. For example, \citet{doak2008} reviewed a collection of field
experiment population dynamic surprises and conclude that these are more common
than we typically assume. \citet{nunez2012} compiled a list of major events on
geologic and contemporary scales that could be considered ecological black
swans. Other work has identified heavy-tail dynamics for particular systems.
For example, \citet{segura2013} identified heavy tails in time series of marine
microbe populations and \citet{drake2014} identified heavy tails in time to
extinction for experimental waterflea populations. However, systematic evidence
has been limited. \citet{keitt1998} tested for power-law tail distributions in
breeding bird abundance, but later worked showed the apparent heavy tails were
a result of mixing populations of different species \citep{allen2001}. A later
large scale analysis of the Global Population Dynamics Database (GPDD) by
\citet{halley2002} found little evidence for heavy-tail distributions in longer
abundance time series.

@@

Our main analysis focuses on the commonly applied phenomenological Gompertz
population dynamics model. In previous analyses of the GPDD, the Gompertz has
proved to be the most frequently parsimonious model \citep{brook2006}. We
extended the traditional Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$
(Fig.~\ref{fig:didactic}). If $\nu$ is small ($\lesssim 10$) the t distribution
has much heavier tails than a normal distribution. For example, at $\nu = 2$,
the probability of drawing a value more than five standard deviations below the
mean is $1.8$\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity
the distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

The heavy-tailed Gompertz model is:
\begin{align*}
x_t &= \ln N_t\\
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align*}
where $x_t$ is the $\ln$ abundance ($N$) at time $t$. The model is density
independent if $b = 1$, maximally density dependent if $b = 0$, and inversely
density dependent if $b < 0$. The parameter $\lambda$ represents the expected
$\ln$ number of new individuals at time $t + 1$ if $N_t = 1$ (per capita
population growth rate). The process noise is modelled as a Student-t
distribution with degrees of freedom parameter $\nu$, a mean of $0$, and
a scale parameter of $\sigma$.

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics \citep[Supporting Material]{gelman2014}. For
$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
values above two (Fig.~\ref{fig:priors}), a slightly less informative prior
than that suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution that time series of the
length considered here are unlikely to be informative about the precise value
of $\nu$ (Fig.~\ref{fig:didactic}).

We estimated all models in a Bayesian framework using Stan
\citep{stan-manual2014}. Stan samples from the posterior distribution with
a version of Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

We fit four alternative population models to test if they systematically
changed our conclusions. (1) Autocorrelation has been suggested as a reason for
increased observed variability of abundance time series through time, which
could create apparent heavy tails \citep{inchausti2002}; therefore, we fit
a model that modelled the serial correlation in the residuals. (2) Previous
work has focussed modelled growth rates without accounting for density
dependence \citep{segura2013}; therefore, we fit a simpler model in which we
assumed density independence. (3) Observation error could bias parameter
estimates \citep{knape2012} or mask our ability to detect heavy tails
\citep{ward2007}. However, simultaneously estimating observation error and
process noise in a state space framework is a challenging computational problem
\citep[e.g.][]{dennis2006, knape2008} even with longer time series and without
estimating the shape of the process error tails; therefore, we fit a model
where we assumed a level of observation error ($0.2$ standard deviations on
a log scale). (4) The Gompertz model assumes that population growth rate
declines linearly with log abundance. We also fit an alternative model, the
Ricker-logistic model, which assumes that population growth rate declines
linearly with abundance itself (Supporting Material).

@@

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
as a random walk with drift, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of
XX populations although the majority of $\nu$ estimates remained qualitatively
similar (Fig.~\ref{fig:alt})

@@

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag in
the UK was thought to reduce population productivity and make the population
vulnerable when a red-tide hit in 1968. The population experienced a black swan
downswing, but the new availability of nest sites resulted in a rapid
population upswing \citep{potts1980}. As another example, an interaction
between population cycles caused by predation combined with cycles caused by
the environment are thought to have caused a downward black swan for a water
vole population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For example, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds that
the last event was a combination of two black swan events in a row and
population recovery took three times longer than predicted
\citep{stafford1971}.

@@

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We
detected black swans more frequently for longer time series and for populations
with a smaller magnitude of process error, as we might expect. However, we
failed to find strong evidence that black swans were associated with density
dependence, population growth rate, or lifespan. In cases that we verified,
black swans were often a result of the interaction between elements of extreme
climate, predators and parasite cycles, and strong changes in human pressures.
Our empirical results, sensitivity tests, and simulation tests suggest that
estimating the tail shape of process error is a viable method of detecting
heavy-tailed population dynamics. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond to,
and are robust to extremes in population dynamics --- particularly as the
frequency and magnitude of climatic extremes increase over the next century
(REF).

@@

We might expect to observe black swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions, we could observe heavy tails if we miss some underlying
mixture of intrinsic processes \citep{allen2001}. That process might be an
aggregation of populations across space, or population diversity, or some
intrinsic change in population variability through time. Extrinsic forces could
also cause black swan dynamics (REF). These forces could be extreme themselves.
For example, extreme climate, predation from or competition with other species
experiencing black swans, or sharp changes in human pressure such hunting,
fishing, or habitat destruction might cause black swans (REFS). Alternatively,
the interaction of multiple ``normal'' extrinsic forces could give rise to
black swan ecological dynamics. This could occur if the interaction nature is
synergistic \citep[e.g.][]{kirby2009} or even if non-synergistic interactions
experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point. Recording
mistakes undoubtedly comprise some of the black swans we detected. Conversely,
recording mistakes could mask heavy tails if initial observations assumed to be
recording errors and are discarded or altered. Third, the temporal scale of
observation and population dynamics vary considerably across populations in the
GPDD and this likely influences the detection of heavy tails. For example, if
we make frequent observations relative to generation time (e.g.~for many large
bodied mammals) we will average across generations and perhaps not observe
black swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans can and do occur, can we forecast the probability
of black swans in space and time? Furthermore, what management policies allow
us to detect them quickly after they happen? Secondly, can we isolate the
components of ecological dynamics that experience black swans events by moving
from phenomenological models such as the Gompertz to mechanistic models, that
take into account, for example, recruitment? We expect that greater insight
into the mechanisms and covariates of ecological black swans may be best
obtained through specific geographic and taxonomic subsets of data where longer
time series are available \citep[e.g.][]{segura2013}. Finally, what is the
impact of allowing for black swans in forecasts of ecological risk? Financial
analysts are concerned with the shape of financial-return distribution tails
because these impact estimates of risk --- the probability of a specific
magnitude of undesired event occurring (REF). It is possible that a comparable
focus in ecology would alter our perception of ecological forecasts such as
extinction risk.

@@

unlikely to forecast black swans, at least with any useful advanced warning
but we might be able to determine probabilities
and develop policy that is robust to, encourages rapid detection, and that is flexible enought to respond quickly to black swans

example of earch quakes - spatial planning, early warning systems, education systems, a general idea about the probability in given places, build buildings that are robust

@@

We thank members of the Earth to Ocean research group for helpful discussions
and comments. We are grateful to the contributors and maintainers of the Global
Population Dynamics Database and to Compute Canada's WestGrid high-performance
computing resources. Three silhouette images were obtained from
\texttt{phylopic.org} under a Creative Commons Attribution 3.0 Unported
license: rabbit (Sarah Werning), hoverfly (Gareth Monger), bird (Jean-Raphaël
Guillaumin {[}photography{]} and T. Michael Keesey {[}vectorization{]}).
Funding was provided by a Simon Fraser University Graduate Fellowship (SCA),
the Natural Sciences and Engineering Research Council of Canada (NKD, ABC), the
Canada Research Chairs Program (NKD), and XX-TODO (TAB).

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamics models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value less than -5 is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
dynamics from a Gompertz model with process noise drawn from t distributions
with different values of $\nu$. Coloured dots in panels c and d represent jumps
with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
are the posterior samples (histograms), median and interquartile range of the
posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
(dashed lines). Colour shading behind panels f--h illustrates the region of
heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
populations of birds, mammals, insects, and fishes. Small values of $\nu$
($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
approximately normal-tailed dynamics. Vertical points and line segments
represent posterior medians and 50\% / 90\% credible intervals for individual
populations. Inset plots show probability that $\nu < 10$ (probability of heavy
tails) for populations arranged by taxonomic order. Taxonomic orders are sorted
by decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
populations in panel a are omitted for space. Red and orange colours highlight
populations with a high probability of heavy-tailed dynamics.
}
\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential correlates of heavy-tailed behaviour (indicated by a high probability
that $\nu < 10$). Shown are (a--c) parameters from the Gompertz heavy-tailed
population model, (b) number of time steps, (c) body length, and (d) lifespan.
For the Gompertz parameters, $\sigma$ refers to the scale parameter of the
Student-t process-noise distribution, $\lambda$ refers to the population growth
rate at an abundance of one, $b$ refers to the density dependence parameter
($1$ is maximally density independent, $0$ is maximally density dependent, and
$<0$ is inversely density dependent).
%For parameter values, points and lines represent medians and 50\% credible
%intervals of the posterior.
Circles representing a limited number of sharks, crustaceans, and gastropods
are filled in white. Median and 90\% credible interval posterior predictions of
a beta regression multilevel model are shown in panels a and d where there was
a high probability the slope coefficient was different from zero (Supporting
Material).
}
\label{fig:correlates}
\end{center}
\end{figure}

>>>>

%Existing evidence for black swans or heavy tails in population dynamics is
%mostly anecdotal or for particular systems.
There is compelling anecdotal evidence for ecological black swans, and
quantitative evidence for a limited number of systems, but systematic evidence
has been elusive. \citet{doak2008} reviewed a collection of field experiment
population dynamic surprises and conclude that these are more common than we
typically assume. \citet{nunez2012} compiled a list of major events on geologic
and contemporary scales that could be considered ecological black swans. Other
work has identified heavy-tail dynamics for particular systems. For example,
\citet{segura2013} identified heavy tails in time series of marine microbe
populations and \citet{drake2014} identified heavy tails in time to extinction
for experimental waterflea populations. However, systematic evidence has been
limited. \citet{keitt1998} tested for power-law tail distributions in breeding
bird abundance, but later worked showed the apparent heavy tails were a result
of mixing populations of different species \citep{allen2001}. A later large
scale analysis of the Global Population Dynamics Database (GPDD) by
\citet{halley2002} found little evidence for heavy-tail distributions in longer
abundance time series.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model (REFs?). The Gompertz model represents population growth as a linear
function in $\log$ space. If we let $x_t$ represent the $\log$ abundance ($N$)
at time $t$, we can represent the Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious model
\citep{brook2006}.

We extended the standard Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$, and scale parameter $\sigma$:
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution. For example, at $\nu = 2$, the probability of
drawing a value more than five standard deviations below the mean is $1.8$\%,
whereas the probability of drawing such a value from a normal distribution is
nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

We chose weakly informative priors to incorporate our understanding of
plausible population dynamics \citep[Supporting Material]{gelman2014}. For
$\nu$, we chose an exponential prior with rate parameter of $0.01$ truncated at
values above two (Fig.~\ref{fig:priors}) --- a slightly less informative prior
than suggested by \citet{fernandez1998}. This prior gives only a 7.7\%
probability that $\nu < 10$ but constrains the sampling sufficiently to avoid
wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution that time series of the
length considered here are unlikely to be informative about the precise value
of $\nu$ (Fig.~\ref{fig:didactic}).

We estimated all models in a Bayesian framework using Stan
\citep[][version 2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of
Hamiltonian Markov chain Monte Carlo called the No-U-Turn Sampler and
generally obtains less correlated chains than algorithms such as the Gibbs
sampler \citep{hoffman2014}. We assured that chains had sufficiently converged
and the sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

We fit four alternative population models to test if they systematically
changed our conclusions. (1) Autocorrelation has been suggested as a reason for
increased observed variability of abundance time series through time, which
could create apparent heavy tails \citep{inchausti2002}; therefore, we fit
a model that modelled the serial correlation in the residuals. (2) Previous
work has modelled abundance or growth rates without accounting for density
dependence \citep{halley2002,segura2013}; therefore, we fit a simpler model in
which we assumed density independence. (3) Observation error could bias
parameter estimates \citep{knape2012} or mask our ability to detect heavy tails
\citep{ward2007}; therefore, we fit a model where we allowed for a fixed
quantity of observation error ($0.2$ standard deviations on a log scale). (4)
The Gompertz model assumes that population growth rate declines linearly with
log abundance. We also fit an alternative model, the Ricker-logistic model,
which assumes that population growth rate declines linearly with abundance
itself (Supporting Material).

% TODO - make sure this is in the supp? However, simultaneously estimating
% observation error
%and process noise in a state space framework is a challenging computational
%problem \citep[e.g.][]{dennis2006, knape2008} even with longer time series and
%without estimating the shape of the process error tails;

@@

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
without density dependence, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in X of
XX populations although the majority of $\nu$ estimates remained qualitatively
similar (Fig.~\ref{fig:alt})

@@

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag
long-necked bird in the UK was thought to reduce population productivity and
make the population vulnerable when a red-tide hit in 1968 \citep{potts1980}.
The population experienced a black swan downswing, but the new availability of
nest sites resulted in a rapid population upswing \citep{potts1980}. As
another example, an interaction between environmentally and predation-mediated
population cycles are thought to have caused a downward black swan for a water
vole population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For example, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds
that the last event was a combination of two black swan events in a row and it
took the population three times longer to recover than predicted
\citep{stafford1971}.

@@

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in XX--XX\% of populations assessed. We
detected black swans more frequently for longer time series and for populations
with a smaller magnitude of process error, as we might expect. However, we
failed to find strong evidence that black swans were associated with density
dependence, population growth rate, or lifespan. In cases that we verified,
black swans were often a result of the interaction between elements of extreme
climate, predators and parasite cycles, and strong changes in human pressures.
Our empirical results, sensitivity tests, and simulation tests suggest that
estimating the tail shape of process error is a viable method of detecting
heavy-tailed population dynamics and if anything will underestimate the
probability of black swans. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond
to, and are robust to extremes in population dynamics --- particularly as the
frequency and magnitude of climatic extremes increase over the next century
(REF).

@@

We might expect to observe black swan dynamics in ecological time series
because of unmodelled intrinsic properties of populations or extrinsic forces
acting on populations. Since a t-distribution can be formed by a mixture of
normal distributions, we could observe heavy tails if we miss some underlying
mixture of intrinsic processes \citep{allen2001}. That process might be an
aggregation of populations across space, or population diversity, or some
intrinsic change in population variability through time. Extrinsic forces
could also cause black swan dynamics \citep[e.g.][]{nunez2012}. These forces
could be extreme themselves. For example, extreme climate, predation from or
competition with other species experiencing black swans, or sharp changes in
human pressure such hunting, fishing, or habitat destruction might cause black
swans. Alternatively, the interaction of multiple ``normal'' extrinsic
forces could give rise to black swan ecological dynamics. This could occur if
the interaction nature is synergistic \citep[e.g.][]{kirby2009} or even if
non-synergistic interactions experience a rare alignment \citep{denny2009}.

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point. Recording
mistakes undoubtedly comprise some of the black swans we detected. Conversely,
recording mistakes could mask heavy tails if initial observations are assumed
to be recording errors and are discarded or altered. Third, the temporal scale
of observation and population dynamics vary considerably across populations in
the GPDD and this likely influences the detection of heavy tails. For example,
if we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss
black swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans can and do occur, can we forecast the probability
of black swans in space and time? Furthermore, what management policies allow
us to detect them quickly after they happen? Secondly, can we isolate the
components of ecological dynamics that experience black swans events by moving
from phenomenological models such as the Gompertz to mechanistic models, that
take into account, for example, recruitment? We expect that greater insight
into the mechanisms and covariates of ecological black swans may be best
obtained through specific geographic and taxonomic subsets of data where longer
time series are available \citep[e.g.][]{segura2013}. Finally, what is the
impact of allowing for black swans in forecasts of ecological risk? Financial
analysts are concerned with the shape of financial-return distribution tails
because these impact estimates of risk --- the probability of a specific
magnitude of undesired event occurring \citep{rachev2008}. It is possible that
a comparable focus in ecology would alter our perception of ecological
forecasts such as extinction risk.

@@

unlikely to forecast black swans, at least with any useful advanced warning
but we might be able to determine probabilities and develop policy that is
robust to, encourages rapid detection, and that is flexible enough to respond
quickly to black swans

example of earth quakes - spatial planning, early warning systems, education
systems, a general idea about the probability in given places, build buildings
that are robust

@@

We thank Justin D. Yeakel, TODO, and other members of the Earth to Ocean
research group for helpful discussions and comments. We are grateful to the
contributors and maintainers of the Global Population Dynamics Database and to
Compute Canada's WestGrid high-performance computing resources. Three
silhouette images were obtained from \texttt{phylopic.org} under a Creative
Commons Attribution 3.0 Unported license: rabbit (Sarah Werning), hoverfly
(Gareth Monger), bird (Jean-Raphaël Guillaumin {[}photography{]} and T.
Michael Keesey {[}vectorization{]}). Funding was provided by a Simon Fraser
University Graduate Fellowship (SCA), the Natural Sciences and Engineering
Research Council of Canada (NKD, ABC), the Canada Research Chairs Program
(NKD), and TODO (TAB).

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value less than -5 is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
dynamics from a Gompertz model with process noise drawn from t distributions
with different values of $\nu$. Coloured dots in panels c and d represent jumps
with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
are the posterior samples (histograms), median and interquartile range of the
posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
(dashed lines). Colour shading behind panels f--h illustrates the region of
heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
populations of birds, mammals, insects, and fishes. Small values of $\nu$
($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
approximately normal-tailed dynamics. Vertical points and line segments
represent posterior medians and 50\% / 90\% credible intervals for individual
populations. Inset plots show probability that $\nu < 10$ (probability of
heavy tails) for populations arranged by taxonomic order and sorted by
decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
populations in panel a are omitted for space. Red to yellow points highlight
populations with a high probability of heavy-tailed dynamics.
} \label{fig:nu-coefs} \end{center} \end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/correlates-p10.pdf}
\caption{
Potential covariates of heavy-tailed population dynamics (indicated by a high
probability that $\nu < 10$). Shown are (a--c) parameters from the Gompertz
heavy-tailed population model, (b) number of time steps, (c) body length, and
(d) lifespan. For the Gompertz parameters, $\sigma$ refers to the scale
parameter of the Student-t process-noise distribution, $\lambda$ refers to the
expected log abundance at the next time step at an abundance of one, $b$
refers to the density dependence parameter ($1$ is maximally density
independent, $0$ is maximally density dependent, and $<0$ is inversely density
dependent).
%For parameter values, points and lines represent medians and 50\% credible
%intervals of the posterior.
Circles representing a limited number of sharks, crustaceans, and gastropods
are filled in white. Median and 90\% credible interval posterior predictions of
a beta regression multilevel model are shown in panels a and d where there was
a high probability the slope coefficient was different from zero (Supporting
Material).
}
\label{fig:correlates}
\end{center}
\end{figure}
